\bf Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2016-02-02 jasa/paper-jasa.tex
Many small changes

The likelihood function is the central object that summarizes the information
from an experiment needed for inference of model parameters. In particular, it
is key to many areas of science that report the results of classical
hypothesis tests or confidence intervals using the (generalized or profile)
likelihood ratio as a test statistic. At the same time, with the advance of
computing technology, it has become increasingly common that a simulator (or
generative model) is used to describe complex processes that tie parameters
$\theta$ of an underlying theory and measurement apparatus to high-dimensional
observations $\mathbf{x}$. However, directly evaluating the likelihood function
in these cases is often impossible or is computationally impractical.

@@

% p_U(u=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
p_U(u=s(\mathbf{x})|\theta_0) &= \frac{d}{du}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq u\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\

@@

p_U(u=s(\mathbf{x})|\theta_0) = \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}

@@

% p_U(u=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = y\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \nonumber \\
p_U(u=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\

@@

Similarly, Equation~\ref{eqn:hormander} can be used to derive $p_U(u=s(\mathbf{x})|\theta_1)$, finally yielding

@@

\frac{p_U(u=s(\mathbf{x})|\theta_0)}{p_U(u=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}.

@@

In light of this result, the likelihood ratio estimation problem can now
be recast as a (probabilistic) classification problem, by noticing that the regression
function
\begin{equation}\label{eqn:best-s-clf}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies conditions of
Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
In other words, supervised learning yields a sufficient
procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{1}{1 +
r(\mathbf{x})}$ only.
Equivalently,
Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
$p_U(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.

@@

As outlined above, any consistent probalilistic classification algorithm can be
used for learning an approximate  map $\hat{s}(\mathbf{x})$ of
Eqn.~\ref{eqn:best-s-clf}. In the common case where the density ratio is
expected to smoothly vary around $\mathbf{x}$, we would however recommend
learning models whose output value $\hat{s}(\mathbf{x})$ also smoothly varies
around $\mathbf{x}$, such as neural networks. In particular, tree-based
methods are not expected to work so well for this use case, since they usually
model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
\glnote{We should comment as to what happens if $\hat{s}(x)$ is not monotonic with $r$}

Given a reduction map $s$, our results show that a statistic equivalent to the
likelihood ratio can be constructed, provided $p(s(\mathbf{x})|\theta)$ can be
evaluated. Again, we do not have a direct and exact way for evaluating this
density, but an approximation $\hat p(\hat s(\mathbf{x})| \theta)$ can be built
instead, e.g. using density estimation or calibration algorithms, such as
histograms, KDE or isotonic regression, as applied on $\{\hat
s(\mathbf{x}_i)\}$, for samples $\{\mathbf{x}_i\}$  drawn from the generative
model. Most notably, learning such an approximation of $p(s(\mathbf{x})|\theta)$
is a much simpler problem than learning $p(\mathbf{x}|\theta)$, since the
reduction $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
the (simpler) informative content of $r(\mathbf{x})$ is preserved.

One strength of the proposed approach is that it factorizes the approximation of the
per-sample reduction ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from
the calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor job at
reproducing the level sets of the per-event likelihood ratio, the density of
$\hat{s}$ can still be well calibrated. In that case, one might loose power, but
the resulting inference will still be valid.
This point was made by \cite{Neal:2007zz} and is well appreciated by the
particle physics community that typically takes a conservative attitude towards
the use of machine learning classifiers precisely due to concerns about the
calibration of $p$-values in the face of nuisance parameters associated to the
simulator.

\glnote{Add pseudo-code to summarize the whole pipeline?}

@@

In order to provide parameter inference in the likelihood-free setting as
described above, we must train a family $s(\mathbf{x};\theta,\theta_1)$ of classifiers parameterized by
$\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
hypotheses, respectively. While this could be done independently for all
$\theta_0$ and $\theta_1$, using the procedure outlined in
Section~\ref{sec:likelihood-ratio-tests}, it is desirable and convenient to have a
smooth evolution of the classification score as a function of the parameters.
For this reason, we anticipate a single learning stage based on training data with input
$(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
unknown values of the parameters are taken as input to the classifier; their
values will be specified via the enveloping (generalized) likelihood ratio of
Eqn.~\ref{eqn:generalized-lr}.

@@

While the function
$p(\mathbf{x}|\theta_1)/(p(\mathbf{x}|\theta_0)+p(\mathbf{x}|\theta_1))$ will
minimize the expected squared error loss based on training data produced
according to Algorithm~\ref{alg:training}, it is not clear how training data
from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real
world classifier with finite capacity. This is left as an area for future work.

\glnote{Shall we keep Algorithm~\ref{alg:training} or rather give a sketch in
words? In its current state, the algorithm does not explain how to build ${\cal
L}$ when the parameter spaces are continuous. It also does not specify how many
samples should be generated per parameter value.}

@@

Once the classifier is trained, we can use the generative model together with a
univariate density estimation technique (e.g. histograms or kernel density
estimation) to approximate $p(s|\theta)$ for specific parameter
points. For a single parameter point, this is a tractable univariate density
estimation problem. The challenge comes from the need to calibrate this density
for all values of $\theta$. A straight forward approach would be to run the
generative model on demand for any particular value of $\theta$. In the context
of a likelihood fit this would mean that the optimization algorithm that is
trying to maximize the likelihood with respect to $\theta$ needs access to the
generative model $p(\mathbf{x}|\theta)$. This can be  impractical when the generative
model is computationally expensive or has high-latency (for instance some human
intervention is required to reconfigure the generative model).  In HEP, with a
fixed classifier, it has become common  to interpolate the distribution between
discrete values of $\theta$ in order to produce a continuous parameterization for
$p(s | \theta)$~\citep{Cranmer:2012sba}.
One can easily imagine a number of approaches to embedding the classifier and
estimating the density $p(s|\theta)$ and the relative merits of those
approaches will depend critically on the dimensionality of $\theta$ and the
computational cost of the generative model. We leave a more general strategy for
this overarching optimization problem as an area of future work.

@@

In the special case of (simple or composite) hypothesis testing between
models defined as mixtures of several components, i.e. when
\begin{equation}
p(\mathbf{x}|\theta)=\sum_c w_c(\theta) p_c(\mathbf{x}| \theta),
\end{equation}
the target likelihood ratio can be rewritten in terms of pairwise
classification problems. We indeed have
\begin{align}
\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)} &= \frac{\sum_c w_c(\theta_0) p_c(\mathbf{x}| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(\mathbf{x}| \theta_1)} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(\mathbf{x}| \theta_1)}{  p_c(\mathbf{x}| \theta_0)}  \right]^{-1} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_1)}{ p_c(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_0)}  \right]^{-1}. \label{eq:decomposedResult}
\end{align}
The second line is a trivial, but a useful decomposition into pairwise
density ratio sub-problems between $p_{c'}(\mathbf{x}|\theta_1)$ and
$p_c(\mathbf{x}|\theta_0)$.  The third line uses
Theorem~\ref{thm:ratio-equivalence} to relate the high-dimensional likelihood
ratio into an equivalent calibrated likelihood ratio based on the univariate
density of the corresponding classifier.

In applications where mixture models are commonly used, e.g. as in high energy
physics, this decomposition allows one to construct better likelihood ratio
estimates since it forces classifiers $s_{c,c'}$ to focus on simpler
sub-problems, for which higher accuracy is expected.

@@

More generally, likelihood ratio testing directly relates to the density ratio
estimation problem, which consists in estimating the ratio of two probability
densities from given and finite collections of observations ${\cal D}_0$ and
${\cal D}_1$. Density ratio estimation is connected to many machine learning
fundamental problems, including transfer learning~\citep{sugiyama2012machine},
probabilistic classification and regression~\citep{vapnik1998statistical},
outlier detection~\citep{hido2011statistical}, and many others. For learning
under covariate shift, \cite{shimodaira2000improving} and \cite{sugiyama2005input} estimate
the density ratio $r(\mathbf{x};\theta_0,\theta_1) =
\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ from straightforward
approximations $\hat{p}(\mathbf{x}|\theta_0)$ and $\hat{p}(\mathbf{x}|\theta_1)$
separately obtained using kernel density estimation. Despite its theoretical
consistency, this approach is known to be ineffective in
practice~\citep{sugiyama2007covariate,bickel2009discriminative}, in particular
as it relies on modeling numerator and denominator high-dimensional densities,
which is a harder problem than modeling their ratio only.
While the proposed method also proceeds in two similar steps, estimating
$p(s(\mathbf{x}))$ is a much simpler problem than estimating $p(\mathbf{x})$,
since $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
the informative content of $r(\mathbf{x})$ is preserved.
Finally, in contrast with the
proposed method which decouples reduction from calibration, other
approaches proposed within the literature (see
\cite{sugiyama2012density,gretton2009covariate,nguyen2010estimating,vapnik2013constructive}
and references therein) also provide solutions for estimating
$r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$, in one step. Under some
assumptions, the convergence of the obtained estimates is also proven for some
of these approaches.

@@

\glnote{Improve given new draft.}
In this work, we have outlined an approach to reformulate generalized likelihood
ratio testing over a high-dimensional data set in terms of a univariate density
of a classifier score. We have shown that a parameterized family of
discriminative classifiers $\hat s(\mathbf{x}; \theta_0, \theta_1)$ trained and
calibrated with a simulator can be used to approximate the likelihood ratio,
even when it is not possible to directly evaluate the likelihood
$p(\mathbf{x}|\theta)$.
% A technique for decomposing this ratio when the generative model is a mixture
% of components was presented with the aim to help focus capacity of the
% classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small
% mixture coefficient.  This approach leverages the power of machine learning in a
% classical statistical setting.
The proposed method offers an alternative to approximate Bayesian computation
for parameter inference in the likelihood-free setting that can also be used in
the frequentist formalism without specifying a prior over the parameters. A
strength of this approach is that it separates the quality of the approximation
of the target likelihood from the quality of the calibration. The former is
related to the ability of supervised learning approaches to  classification,
which will continue to improve. The calibration procedure for a particular
parameter point is fairly straight forward since it involves estimating a
univariate density using a generative model of the data. The difficulty of the
calibration stage is performing this calibration continuously in $\theta$.
Different strategies to this calibration are anticipated depending on the
dimensionality of $\theta$, the complexity of the resulting likelihood function,
and the difficulties associated to running the simulator.

>>>>

The likelihood function is the central object that summarizes the information
from an experiment needed for inference of model parameters. It
is key to many areas of science that report the results of classical
hypothesis tests or confidence intervals using the (generalized or profile)
likelihood ratio as a test statistic. At the same time, with the advance of
computing technology, it has become increasingly common that a simulator (or
generative model) is used to describe complex processes that tie parameters
$\theta$ of an underlying theory and measurement apparatus to high-dimensional
observations $\mathbf{x}$. However, directly evaluating the likelihood function
in these cases is often impossible or is computationally impractical.

@@

% p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{dy} P(s(\mathbf{X}) \leq y) \nonumber \\
p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \frac{d}{du}  \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) \leq u\}} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) d\mathbf{x}^\prime \nonumber \\

@@

p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) = \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}} \frac{1}{|\nabla s(\mathbf{x}^\prime)|} p_\mathbf{X}(\mathbf{x}^\prime|\theta_0) dS_{\mathbf{x}^\prime} \label{eqn:hormander}

@@

p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0) &= \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|} m^{-1}(s(\mathbf{x}^\prime)) p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} \nonumber \\

@@

Similarly, Equation~\ref{eqn:hormander} can be used to derive $p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)$, finally yielding

@@

\frac{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_0)}{p_{\mathbf{U}}(u=s(\mathbf{x})|\theta_1)} &= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)} \frac{\int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime}}{ \int_{\{\mathbf{x}^\prime \in \mathbb{R}^p : s(\mathbf{x}^\prime) = u\}}  \frac{1}{|\nabla s(\mathbf{x}^\prime)|}  p_\mathbf{X}(\mathbf{x}^\prime|\theta_1) dS_{\mathbf{x}^\prime} } \nonumber \\
&= \frac{p_\mathbf{X}(\mathbf{x}|\theta_0)}{p_\mathbf{X}(\mathbf{x}|\theta_1)}. \label{eqn:jacob}

@@

In light of this result, the likelihood ratio estimation problem can now
be recast as a (probabilistic) classification problem, by noticing that the decision
function
\begin{equation}\label{eqn:best-s-clf}
s^*(\mathbf{x}) = \frac{p_{\mathbf{X}}(\mathbf{x}|\theta_1)}{p_{\mathbf{X}}(\mathbf{x} | \theta_0) + p_{\mathbf{X}}(\mathbf{x} | \theta_1)}.
\end{equation}
modeled by a classifier trained to distinguish samples $\mathbf{x} \sim p_{\theta_0}$
from samples $\mathbf{x} \sim p_{\theta_1}$ satisfies conditions of
Theorem~\ref{thm:ratio-equivalence} (see Appendix~\ref{app:clf-for-s} for further details).
In other words, supervised learning yields a sufficient
procedure for Theorem~\ref{thm:ratio-equivalence} to hold, guaranteeing that any
{\it universally strongly consistent} algorithm can be used for learning $s^*$.
Note however, that it is not a necessary procedure since
Theorem~\ref{thm:ratio-equivalence} holds for any monotonic function $m$ of the
density ratio, i.e., not for $m(r(\mathbf{x})) = \frac{1}{1 +
r(\mathbf{x})}$ only.
Equivalently,
Theorem~\ref{thm:ratio-equivalence} shows that in case we learn a probabilistic
classifier $s(\mathbf{x})$ which is imperfect up to a monotonic transformation
of $r(\mathbf{x})$, then one can still resort to calibration (i.e., modeling
$p_{\mathbf{U}}(u=s(\mathbf{x}))$) to compute $r(\mathbf{x})$ exactly.

@@

As outlined above, any consistent probalilistic classification algorithm can be
used for learning an approximate  map $\hat{s}(\mathbf{x})$ of
Eqn.~\ref{eqn:best-s-clf}. In the common case where the density ratio is
expected to smoothly vary around $\mathbf{x}$, we would however recommend
learning models whose output value $\hat{s}(\mathbf{x})$ also smoothly varies
around $\mathbf{x}$, such as neural networks. For small training sets, tree-based
methods are not expected to work so well for this use case, since they usually
model $\hat{s}(\mathbf{x})$ as a non-strictly monotonic composition of step functions.
In particular, in such cases where $s(\mathbf{x})$ is not monotonic with $r(\mathbf{x})$, Jacobian
factors in Eqn.~\ref{eqn:jacob} do not cancel, resulting in a biased
approximation of the density ratio.

Given a reduction map $s$, our results show that a statistic equivalent to the
likelihood ratio can be constructed, provided $p(s(\mathbf{x})|\theta)$ can be
evaluated. Again, we do not have a direct and exact way for evaluating this
density, but an approximation $\hat p(\hat s(\mathbf{x})| \theta)$ can be built
instead, e.g. using density estimation or calibration algorithms, such as
histograms, KDE or isotonic regression, and as applied on $\{\hat
s(\mathbf{x}_i)\}$, for samples $\{\mathbf{x}_i\}$  drawn from the generative
model. Most notably, learning such an approximation of $p(s(\mathbf{x})|\theta)$
is a much simpler problem than learning $p(\mathbf{x}|\theta)$, since the
reduction $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
the (simpler) informative content of $r(\mathbf{x})$ is preserved.

One strength of the proposed approach is that it factorizes the approximation of
the per-sample reduction ($\hat{s}(\mathbf{x}) \approx s(\mathbf{x})$) from the
calibration procedure ($\hat p(\hat s(\mathbf{x})| \theta) \approx
p(\hat{s}(\mathbf{x})|\theta)$). Thus, even if the classifier does a poor at
learning the optimal decision function~\ref{eqn:best-s-clf} and therefore at
reproducing the level sets of the per-sample likelihood ratio, the density of
$\hat{s}$ can still be well calibrated. In that case, one might loose power, but
the resulting inference will still be valid. This point was made by
\cite{Neal:2007zz} and is well appreciated by the particle physics community
that typically takes a conservative attitude towards the use of machine learning
classifiers precisely due to concerns about the calibration of $p$-values in the
face of nuisance parameters associated to the simulator.

@@

In order to provide parameter inference in the likelihood-free setting as
described above, we must train a family $s(\mathbf{x};\theta_0,\theta_1)$ of classifiers parameterized by
$\theta_0$ and $\theta_1$, the parameters associated to the null and alternate
hypotheses, respectively. While this could be done independently for all
$\theta_0$ and $\theta_1$, using the procedure outlined in
Section~\ref{sec:likelihood-ratio-tests}, it is desirable and convenient to have a
smooth evolution of the classification score as a function of the parameters.
For this reason, we anticipate a single learning stage based on training data with input
$(\mathbf{x}, \theta_0, \theta_1)_i$ and target $y_i$, as outlined in Algorithm~\ref{alg:training}. Somewhat unusually, the
unknown values of the parameters are taken as input to the classifier; their
values will be specified via the enveloping (generalized) likelihood ratio of
Eqn.~\ref{eqn:generalized-lr}.
While the function
$p(\mathbf{x}|\theta_1)/(p(\mathbf{x}|\theta_0)+p(\mathbf{x}|\theta_1))$ will
minimize the expected squared error loss based on training data produced
according to Algorithm~\ref{alg:training}, it is not clear how training data
from $\theta'_0 \ne \theta_0$ and $\theta'_1 \ne \theta_1$ will influence a real
world classifier with finite capacity. This is left as an area for future work.

@@

@@

Once the classifier is trained, we can use the generative model together with a
univariate density estimation technique (e.g. histograms or kernel density
estimation) to approximate $p(s|\theta)$ for specific parameter
points. For a single parameter point $\theta$, this is a tractable univariate density
estimation problem. The challenge comes from the need to calibrate this density
for all values of $\theta$. A straightforward approach would be to run the
generative model on demand for any particular value of $\theta$. In the context
of a likelihood fit this would mean that the optimization algorithm that is
trying to maximize the likelihood with respect to $\theta$ needs access to the
generative model $p(\mathbf{x}|\theta)$. This can be  impractical when the generative
model is computationally expensive or has high-latency (for instance some human
intervention is required to reconfigure the generative model).

In high energy physics, with a
fixed classifier, it has become common  to interpolate the distribution between
discrete values of $\theta$ in order to produce a continuous parameterization for
$p(s | \theta)$~\citep{Cranmer:2012sba}.
One can easily imagine a number of approaches to embedding the classifier and
estimating the density $p(s|\theta)$ and the relative merits of those
approaches will depend critically on the dimensionality of $\theta$ and the
computational cost of the generative model. We leave a more general strategy for
this overarching optimization problem as an area of future work.

@@

In the special case of (simple or composite) hypothesis testing between
models defined as mixtures of several components, i.e. when $p(\mathbf{x}|\theta$ can be written as
\begin{equation}
p(\mathbf{x}|\theta)=\sum_c w_c(\theta) p_c(\mathbf{x}| \theta),
\end{equation}
the target likelihood ratio can be formulated in terms of pairwise
classification problems. We indeed have
\begin{align}
\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)} &= \frac{\sum_c w_c(\theta_0) p_c(\mathbf{x}| \theta_0)}{\sum_{c'} w_{c'}(\theta_1) p_{c'}(\mathbf{x}| \theta_1)} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(\mathbf{x}| \theta_1)}{  p_c(\mathbf{x}| \theta_0)}  \right]^{-1} \nonumber \\
&= \sum_c \left[ \sum_{c'} \frac{ w_{c'}(\theta_1)}{w_c(\theta_0)} \frac{ p_{c'}(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_1)}{ p_c(s_{c,c'}(\mathbf{x};\theta_0, \theta_1)| \theta_0)}  \right]^{-1}. \label{eq:decomposedResult}
\end{align}
The second line is a trivial, but a useful decomposition into pairwise
density ratio sub-problems between $p_{c'}(\mathbf{x}|\theta_1)$ and
$p_c(\mathbf{x}|\theta_0)$.  The third line uses
Theorem~\ref{thm:ratio-equivalence} to relate the high-dimensional likelihood
ratio into an equivalent calibrated likelihood ratio based on the univariate
density of the corresponding classifier.

In applications where mixture models are commonly used this decomposition allows
one to construct better likelihood ratio estimates since it forces classifiers
$s_{c,c'}$ to focus on simpler sub-problems, for which higher accuracy is
expected.

@@

More generally, likelihood ratio testing directly relates to the density ratio
estimation problem, which consists in estimating the ratio of two probability
densities from given and finite collections of observations ${\cal D}_0$ and
${\cal D}_1$. Density ratio estimation is connected to many machine learning
fundamental problems, including transfer learning~\citep{sugiyama2012machine},
probabilistic classification and regression~\citep{vapnik1998statistical},
outlier detection~\citep{hido2011statistical}, and many others. For learning
under covariate shift, \cite{shimodaira2000improving} and \cite{sugiyama2005input} estimate
the density ratio $r(\mathbf{x};\theta_0,\theta_1) =
\frac{p(\mathbf{x}|\theta_0)}{p(\mathbf{x}|\theta_1)}$ from straightforward
approximations $\hat{p}(\mathbf{x}|\theta_0)$ and $\hat{p}(\mathbf{x}|\theta_1)$
separately obtained using kernel density estimation. Despite its theoretical
consistency, this approach is known to be ineffective in
practice~\citep{sugiyama2007covariate,bickel2009discriminative}, since
it relies on modeling numerator and denominator high-dimensional densities,
which is a harder problem than modeling their ratio only.
While the proposed method also proceeds in two similar steps, estimating
$p(s(\mathbf{x}))$ is much easier than estimating $p(\mathbf{x})$,
since $s$ projects $\mathbf{x}$ into a one-dimensional space in which only
the informative content of $r(\mathbf{x})$ is preserved.
Finally, in contrast with the
proposed method which decouples reduction from calibration, other
approaches proposed within the literature (see
\cite{sugiyama2012density,gretton2009covariate,nguyen2010estimating,vapnik2013constructive}
and references therein) provide solutions for estimating
$r(\mathbf{x};\theta_0,\theta_1)$ directly from $\mathbf{x}$, in one step. Under some
assumptions, the convergence of the obtained estimates is also proven for some
of these approaches.

@@

In this work, we have outlined an approach to reformulate generalized likelihood
ratio testing over a high-dimensional data set in terms of a univariate density
of a classifier score. We have shown that a parameterized family of
discriminative classifiers $\hat s(\mathbf{x}; \theta_0, \theta_1)$ trained and
calibrated with a simulator can be used to approximate the likelihood ratio,
even when it is not possible to directly evaluate the likelihood
$p(\mathbf{x}|\theta)$.
% A technique for decomposing this ratio when the generative model is a mixture
% of components was presented with the aim to help focus capacity of the
% classifier when $p(x|\theta_0)$ and $p(x|\theta_1)$ differ primarily by a small
% mixture coefficient.  This approach leverages the power of machine learning in a
% classical statistical setting.
The proposed method offers an alternative to approximate Bayesian computation
for parameter inference in the likelihood-free setting that can also be used in
the frequentist formalism without specifying a prior over the parameters. A
strength of this approach is that it separates the quality of the approximation
of the target likelihood from the quality of the calibration. The former is
related to the ability of supervised learning approaches to  classification,
which will continue to improve. The calibration procedure for a particular
parameter point is fairly straightforward since it involves estimating a
univariate density using a generative model of the data. The difficulty of the
calibration stage is performing this calibration continuously in $\theta$.
Different strategies to this calibration are anticipated depending on the
dimensionality of $\theta$, the complexity of the resulting likelihood function,
or the practical issues associated to running the simulator.
